---
title: "Harmful Effects of Severe Weather"
author: "Andrew Luyt"
date: "24/12/2021"
output: 
    html_document:
        keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
*Immediately after the title, there should be a synopsis which describes and 
summarizes your analysis in at most 10 complete sentences.*

## Data Processing
*There should be a section titled Data Processing which describes (in words and 
code) how the data were loaded into R and processed for analysis. In particular, 
your analysis must start from the raw CSV file containing the data. You cannot 
do any preprocessing outside the document. If preprocessing is time-consuming 
you may consider using the `cache = TRUE` option for certain code chunks.*

```{r libraries}
# data.table is required for one function only, "fread". We won't load the
# entire package because its namespace overlaps some other functions we use.
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
```

```{r load-data, cache=TRUE}
df <- as_tibble(data.table::fread(file = "repdata_data_StormData.csv.bz2"))
df <- rename(df, state_num = STATE__, longitude_e = LONGITUDE_)
names(df) <- tolower(names(df))
# these variables are all 0 and all NA, respectively
df <- select(df, -c(county_end, countyendn))


```


## Results
*There should be a section titled Results in which your results are presented.*

