---
title: "Harmful Effects of Severe Weather, 1950 - 2011"
author: "Andrew Luyt"
output: 
    html_document:
        theme: united
        highlight: pygments
        keep_md: true
        number_sections: true
        toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```
*Most recent update: `r Sys.Date()`*


# Synopsis

This report is a final project for the course 
[Reproducible Research](https://www.coursera.org/learn/reproducible-research) by
Johns Hopkins University.

In this analysis we will report on the top weather-related causes of human death & injury
and property & crop damage. To calculate these sums we will use the NOAA's data
set of weather events and associated damage for events in the USA between
1950 and November 2011. We spend substantial effort correcting manual data
entry inconsistencies in the `evtype` variable, and note that damage caused
by hurricanes is split up into multiple categories, for instance flood and hail.

We find the most damaging weather event to human health is the tornado, and
the most damaging weather events to property and crops are flooding and
droughts, respectively.


# Data Processing

We'll load libraries first.

```{r libraries}
# data.table::fread() is also required. We won't load the entire
# package because its namespace overlaps some other functions we use.
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(kableExtra))  # table formatting tools

# Set global table styling options here by overriding kable()
kable <- function(data, ...) {
    knitr::kable(data, format.args = list(big.mark = ","), ...) %>% 
        kable_styling(full_width=FALSE, position="left")
}
```

Next we load the data and perform some basic data cleaning:

* rename two variables with cumbersome names
* remove two variables which contain no information
  * `county_end` and `countyendn` are all 0 and NA, respectively.
* standardize the values of `evtype` to lowercase with no extra spaces.
  * to correct some probable manual data entry errors

```{r load-data, cache=TRUE}
df <- as_tibble(data.table::fread(file = "repdata_data_StormData.csv.bz2"))
df <- rename(df, state_num = STATE__, longitude_e = LONGITUDE_)
names(df) <- tolower(names(df))
# these variables are all 0 and all NA, respectively
df <- select(df, -c(county_end, countyendn))

# standardize `evtype` to lowercase, removing spaces at beginning and end
# and removing extra internal spaces
df$evtype <- tolower(str_squish(df$evtype))
```

## Correcting `propdmgexp` and `cropdmgexp`

In the raw data, damage amounts are represented in an unusual way. They are 
expressed in scientific notation, with the base and exponent stored in 
separate variables.

The exponents, `propdmgexp` and `cropdmgexp`,  are supposed to
represent a power of 10, which is multiplied by `propdmg` or `cropdmg`
respectively to express a final damage amount. However, they actually
contain unusual character values:

```{r check_exponent_features}
sort(union(unique(df$propdmgexp), unique(df$cropdmgexp)))
```

After some investigation, it appears varying data entry standards or data
entry error has resulted in this state of affairs. A power of 10 will be
assigned to replace each of these symbols.
[*For a more complete discussion, see [this document](https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html).*]

* `empty string`: 0
* `+` or `?`: 1
* `0-9`: 10
* `hH`: 100   "hundred"
* `kK`: 1000  "kilo"
* `mM`: 10^6  "million"
* `bB`: 10^9  "billion"
* `-`: NA 

```{r correct_exponent_features, cache=TRUE}
# Perform the substitution described above
old <- c("-", "?", "+", "0", "1", "2", "3", "4", "5", "6", "7", "8", "B", "H", "K", "M", "")
new <- c(NA,   1,   1,   10,  10,  10,  10,  10,  10,  10,  10,  10, 1e9, 1e2, 1e3, 1e6, 0)

ctmp <- toupper(df$cropdmgexp)
ptmp <- toupper(df$propdmgexp)
cnew <- rep(NA_integer_, length(ctmp)) # new vectors will be integers
pnew <- rep(NA_integer_, length(ptmp))

for (i in 1:length(old)) {
    cnew[ctmp == old[[i]]] <- new[[i]]
    pnew[ptmp == old[[i]]] <- new[[i]]
}

df$cropdmgexp <- cnew
df$propdmgexp <- pnew
rm(ctmp, ptmp, cnew, pnew, old, new, i)

```

## Fixing `evtype`

`evtype` may be the most important variable in our dataset. It is the
variable that categorizes each event as `tornado`, `flood` and so forth.
Later we'll use `evtype` to
aggregate human and financial damage for each type of weather event and it's
important that they reflect the categories we think they do.

In the NOAA documentation there are 48 official `evtype` values, corresponding
to the various weather events they wish to record data for.

In the dataset however, `r length(unique(df$evtype))` unique values exist
after standardizing to lowercase and removing extra whitespace. An examination of the values
shows this to have probably resulted from manual data entry. For example,
here we list types relating to blizzards.

```{r demonstrate_data_entry_problems}
df %>% 
    count(evtype) %>% 
    arrange(desc(n)) %>% 
    filter(str_detect(evtype, "blizzard")) %>% 
    head(n=7) %>% 
    kable(caption = 'Counts of blizzard-related evtypes')
```

We can see that `"blizzard"` labels most of the observations, with a small number of
other observations given similar-sounding labels. These observations should probably be
a part of `"blizzard"`. With over 800 types in the dataset, We'd like to know 
if this example represents a systematic problem. 

### What proportion of `evtypes` are problematic?

We constructed a list of official evtypes from NOAA documentation and
stored it in `evtypes.txt`. We wish to know
**what proportion of records have a non-standard `evtype` value?**

```{r find_proportion_of_bad_types}
types <- tolower(readLines("evtypes.txt")) # a vector of standard evtypes
df %>% filter(! evtype %in% types) %>% nrow() / nrow(df)
```
**About 30% of the data is using some other label for `evtype`.** This is
a significant number and could certainly skew our analysis.
e.g. if one major event type
were split into many smaller types because of typos, it might no longer appear
to be a major cause of damage.

### How many non-standard `evtypes` do we need to fix?

What we don't know yet is how these 30% of records are distributed among 
non-standard `evtypes`.  i.e. Are they evenly distributed amongst the 800
or so incorrect `evtypes` (a worst-case scenario) or do a small number of strings
represent the majority of these problematic observations?

We'll group non-standard observations by their `evtype` and count the
observations. We'll also sort these counts and calculate a cumulative
proportion of the observations.

```{r find_cumulative_proportions}
df %>% 
    filter(! evtype %in% types) %>% 
    count(evtype) %>% 
    arrange(desc(n)) %>%  
    mutate(proportion = n / sum(n), cumulative = cumsum(proportion)) %>% 
    slice_max(n, n=4) %>% kable(caption = "Top four errors in evtype")
```

**The four most common errors represent almost 94% of all errors.**

Let's pick a nice round number of observations as a cutoff value. 
We'll count how many non-standard `evtype` strings label 100 or more observations
in the dataset. As a reminder, there are `r nrow(df)` observations in total.

```{r count_types_to_fix_manually}
fixtypes <- df %>% 
    filter(! evtype %in% types) %>%
    count(evtype) %>% 
    arrange(desc(n)) %>% 
    filter(n >= 100) %>% 
    pull(evtype)
length(fixtypes)
```

This is good news, we won't have to fix over 800 errors. **Fixing
the 32 most common** should ensure a much more accurate analysis. Let's calculate
what proportion of observations will be correct, if we fix `evtypes` with
100 or more observations.

```{r}
df %>% 
    filter(! evtype %in% types) %>% 
    count(evtype) %>% 
    arrange(desc(n)) %>% 
    mutate(proportion = n / sum(n), cumulative = cumsum(proportion)) %>% 
    filter(n > 99, n <=101) %>% kable()
```

**98.4% of our problematic records will be corrected after fixing the top 32.**
Let's visualize the situation by plotting the cumulative proportion of correct
`evtypes`.

```{r visualize_evtype_fixes}
x <- df %>% 
    filter(! evtype %in% types) %>% 
    count(evtype) %>% 
    arrange(desc(n)) %>%  
    mutate(proportion = n / sum(n), cumulative = cumsum(proportion)) %>% 
    head(32) 
tmp <- x$cumulative
names(tmp) <- x$evtype
par(mar = c(9,4,2,2), cex.axis = .8)
barplot(tmp, las=2, ylim=c(0,1),
         main="Cumulative proportion correct, after fixing evtype errors")
abline(h=1, lty=2, col='dodgerblue2', lwd=2.5)
```

Fixing `tstm wind` alone will correct over 80% of the records, and then
we see rapidly diminishing returns.

Since evtype errors represent about 30% of all records, correcting these errors
to about 98.4% accuracy (the top 32 errors) will result in an overall `evtype`
accuracy of about `(1 - 0.3 * (1 - .984)) * 100 ~= ` 
`r round((1 - 0.3 * (1 - .984)) * 100, 1)`% for the entire dataset. Further
improvements would be via vastly diminished returns, so we'll consider this
an acceptable improvement over the original data at about 70% correct.

### Fixing the selected `evtypes`

We will work through these 32 types systematically and assign them
the correct `evtype`, using the NOAA documentation and some common sense to
place them in the correct category.

The rest of the non-standard labels we will leave unchanged, assuming they
have too
small an effect on the analysis to matter. Let's first look at the labels
we'll fix.

```{r show_types_to_fix}
kable(matrix(fixtypes, nrow = 8), caption = "The 32 most important evtypes to correct")
```

Now we'll fix these incorrect labels by reassigning the observation to the
correct `evtype`.

```{r create_types_fixlist}
# map official evtypes to a vector of some incorrect names for that type
fixlist <- list(
    "thunderstorm wind" = c("tstm wind", "thunderstorm winds", "tstm wind/hail", 
                            "wind", "dry microburst"),
    "marine thunderstorm wind" = c("marine tstm wind"),
    "flood" = c("urban/sml stream fld", "flood/flash flood", "urban flood",
                "river flood", "flooding"),
    "flash flood" = c("flash flooding"),
    "coastal flood" = c("coastal flooding"),
    "high wind" = c("high winds"),
    "strong wind" = c("strong winds"),
    "wildfire" = c("wild/forest fire"),
    "winter weather" = c("winter weather/mix"),
    "extreme cold/wind chill" = c("extreme cold", "extreme windchill"),
    "heavy snow" = c("snow", "light snow", "moderate snowfall"),
    "debris flow" = c("landslide"),
    "dense fog" = c("fog"),
    "rip current" = c("rip currents"),
    "excessive heat" = c("unseasonably warm", "record warmth"),
    # from docs: "..caused by any combination of...high astronomical tide..."
    "storm surge/tide" = c("storm surge", "astronomical high tide"),  
    "high surf" = c("heavy surf/high surf"),
    "winter weather" = c("freezing rain"),
    "hurricane (typhoon)" = c("hurricane")
)

tmp <- df$evtype

# replace all the bad evtypes with official evtypes
for (officialtype in names(fixlist)) {
    for (badtype in fixlist[[officialtype]]) {
        tmp[grep(pattern = badtype, x = tmp)] <- officialtype
    }
}
df$evtype <- tmp
rm(tmp)
```

Previously, almost 30% of records had an incorrect `evtype`. Let's recalculate 
that proportion.
```{r how_effective_was_fixing_names}
df %>% filter(! evtype %in% types) %>% nrow() / nrow(df)
```

Now, only 0.3% of records have some unusual `evtype` from data
entry errors, or in other words **`evtype` is now almost 99.7% correct.** Earlier
we estimated we'd see 99.5% correct, so this is reasonable.

#### A final test: dollar amounts

Let's make one more test. It's *possible* these small number of 
observations are actually large in terms of human or financial damage. Let us
compare the sum of fatalities, injuries, property damage, and crop damage of the
remaining uncorrected records to the whole. If our suppositions are correct,
the records with non-standard `evtypes` will be only a small percentage of the total
damage.

```{r check_nonstandard_sums}
results <- df %>% 
    mutate(property = propdmg*propdmgexp, crops = cropdmg*cropdmgexp) %>% 
    group_by("standard_evtypes" = evtype %in% types) %>% 
    summarise(across(c(fatalities, injuries, property, crops),
                     ~sum(., na.rm=TRUE))) %>% 
    t()
results <- results[-1, ]
colnames(results) <- c("sum.non.standard", "sum.standard")
as_tibble(results) %>% 
    mutate(non.standard.proportion = sum.non.standard / (sum.non.standard + sum.standard),
           class = rownames(results)) %>% 
    select(class, everything()) %>% 
    kable(caption = "Proportions of total damage represented by non-standard evtypes")
```

With these non-standard cases making up at most 3.3% of the total damage, the dataset
is in much better condition than it was in raw form.
We'll continue the analysis with this imperfect but much-improved dataset, 
assuming that this small proportion of error won't significantly skew our results.

# Results

## Across the United States, which types of events are most harmful to population health?


Here we sum the fatalities and injuries across `evtype`, select the ten largest
sums, and then plot the results.

```{r injuries_fatalities, cache=TRUE}
top_fatalities <- df %>% 
    group_by(evtype) %>% 
    summarise(fatalities = sum(fatalities, na.rm = TRUE)) %>% 
    arrange(desc(fatalities)) %>% 
    slice_head(n = 10)

top_injuries <- df %>% 
    group_by(evtype) %>% 
    summarise(injuries = sum(injuries, na.rm = TRUE)) %>% 
    arrange(desc(injuries)) %>% 
    slice_head(n = 10)
```
```{r graph_injuries_fatalities, fig.height=5, fig.width=7}
par(mfrow = c(1,2), mar=c(c(8, 4, 4, 2) + 0.1), cex.axis=.9)
barplot(top_fatalities$fatalities, names.arg = top_fatalities$evtype, las=2,
        main="Fatalities: top causes")
mtext(text = "Data source: U.S. NOAA storm database, 1950-2011", outer = T,
      line = -25, cex = .7)
barplot(top_injuries$injuries, names.arg = top_injuries$evtype, las=2,
        main="Injuries: top causes")
```

In terms of both fatalities and injuries, **tornadoes are the most destructive
events to human health** in this dataset. 

```{r table_top_fatality_injury}
kable(list(data.frame(rank = 1:10), top_fatalities, data.frame(),data.frame(), top_injuries),
      caption = "Top causes: fatalities & injuries")
```


### Hurricanes?

There appears to be something missing in the above graphs. Knowing something of 
destructive US weather events one must ask **"Where are the hurricanes?"**

There is an important note in the FAQ:

>The fatalities, injuries, and damage amounts appearing in tropical cyclone events are
attributed only to wind damage experienced in the coastal counties/parishes listed. Other
tropical cyclone related events such as tornadoes and flooding are listed within their
separate event types.

**i.e. the deaths and damage from hurricanes have deliberately been split up into 
multiple evtypes** and will show up as part of floods, lightning, wind, *etc.*
Hurricane-spawned tornadoes are also included in the tornado events graphed above.
From the point of view of this dataset, hurricanes may be understood as a
'meta-event' that *causes* other events.

### Why is there both `heat` and `excessive heat`?

It is also worth noting that some similar events have different official
`evtypes`, e.g. `excessive heat` and `heat`, both of which show up in
the graphs above.  We examined the NOAA documentation that accompanies the
dataset and verified that they are indeed separate, official categories. 
*We will leave the situation as is*, assuming the NOAA has good reason to
organize their data thus.

## Across the United States, which types of events have the greatest economic consequences?

We proceed as earlier, summing property and crop damage across all evtypes.

```{r calculate_damages, fig.height=5, fig.width=7}
top_property_dmg <- df %>% 
    group_by(evtype) %>% 
    summarise(property = sum(propdmg * propdmgexp, na.rm = TRUE)) %>% 
    arrange(desc(property)) %>% 
    slice_head(n = 10)
top_crop_dmg <- df %>% 
    group_by(evtype) %>% 
    summarise(crop = sum(cropdmg * cropdmgexp, na.rm = TRUE)) %>% 
    arrange(desc(crop)) %>% 
    slice_head(n = 10)

par(mfrow=c(1,2), mar=c(c(9, 5, 4, 2) + 0.1), cex.axis=.9, mgp = c(4, 1, 0))
barplot(top_property_dmg$property, names.arg = top_property_dmg$evtype, las=2,
        main="Property damage: top causes", ylab = "US Dollars")
mtext(text = "Data source: U.S. NOAA storm database, 1950-2011", outer = T,
      line = -25, cex = .7)
barplot(top_crop_dmg$crop, names.arg = top_crop_dmg$evtype, las=2,
        main="Crop damage: top causes")
```

```{r table_top_damage}
kable(list(data.frame(rank = 1:10), top_property_dmg, data.frame(),data.frame(), 
           top_crop_dmg),
      caption = "Top causes: property & crop damage")
```

As mentioned in the previous section, the amounts for `hurricane (typhoon)`
are **only** for wind damage on coastal communities. Hurricane-caused flooding,
*etc.* are grouped into separate `evtypes`.

<hr noshade />
[author website: andrewluyt.github.io](https://andrewluyt.github.io/)
<hr noshade/>
To aid reproducibility, here is the session environment at time of publishing.
```{r}
sessionInfo()
```

